---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# truncpois

<!-- badges: start -->

<!-- badges: end -->

## Truncated Poisson Distribution Functions

### Written by Arun Sundar Paulraj and Keefe Murphy

## Description

`truncpois` provides a complete suite of distribution functions for the truncated Poisson distribution, supporting both left-truncation, right-truncation, and doubly truncated forms. Truncation bounds are interpreted as `a <= X <= b`. All core computations are carried out on the log scale for numerical stability, making the package reliable across a wide range of rate parameters and truncation configurations.

The most important functions in the __truncpois__ package are: `dtruncpois`, for computing the probability mass function; `ptruncpois`, for the cumulative distribution function; `qtruncpois`, for quantiles; and `rtruncpois`, for generating random samples via inverse-CDF or bounded rejection sampling. Distributional moments are provided by `extruncpois` and `vartruncpois` for the theoretical mean and variance respectively.

## Installation

You can install the latest stable official release of `truncpois` from CRAN:

```r
install.packages("truncpois")
```

or the development version from GitHub:

```r
# install.packages("pak")
pak::pak("arunsundar022/truncpois")
```

In either case, you can then explore the package with:

```r
library(truncpois)
help(dtruncpois)  # Help on the core density function
```

For a more thorough introduction, a vignette document is available as follows:

```r
vignette("truncpois", package = "truncpois")
```

However, if the package is installed from GitHub, the vignette is not automatically built. It can be included when installing from GitHub with:

```r
devtools::install_github("arunsundar022/truncpois", build_vignettes = TRUE)
```

Alternatively, the vignette is available on the package's CRAN page.

## Quick Example

```{r library}
library(truncpois)
set.seed(123)
```

Compare the standard Poisson and right-truncated Poisson PMFs:

```{r pmf-compare}
x <- 0:12
p_pois  <- dpois(x, lambda = 5)
p_trunc <- dtruncpois(x, lambda = 5, a = 0, b = 8)

round(data.frame(x = x, p_pois = p_pois, p_trunc = p_trunc), 4)
```

Generate random draws from a doubly truncated Poisson distribution:

```{r rtruncpois}
samples <- rtruncpois(1000, lambda = 4, a = 2, b = 9, method = "bounded")
table(samples)
```

Compare empirical and theoretical moments:

```{r moments}
c(empirical_mean    = mean(samples),
  theoretical_mean  = extruncpois(4, a = 2, b = 9))

c(empirical_var     = var(samples),
  theoretical_var   = vartruncpois(4, a = 2, b = 9))
```

## References

Nadarajah, S. and Kotz, S. (2006). R programs for truncated distributions.
_Journal of Statistical Software, Code Snippets_, 16(2), 1--8.
<[doi:10.18637/jss.v016.c02](https://doi.org/10.18637/jss.v016.c02)>
