---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Load the package for examples
devtools::load_all()
```

# truncpois

<!-- badges: start -->

<!-- badges: end -->

## Truncated Poisson Distribution Functions

### Written by Arun Sundar Paulraj and Keefe Murphy

## Description

`truncpois` provides a complete suite of distribution functions for the truncated Poisson distribution, supporting left-truncation, right-truncation, and doubly-truncated forms. Truncation bounds are interpreted as `a <= X <= b`. All computations are carried out on the log scale for numerical stability.

The core functions are `dtruncpois()` (density), `ptruncpois()` (CDF), `qtruncpois()` (quantiles), and `rtruncpois()` (random generation). Alongside these, `extruncpois()`, `vartruncpois()`, `medtruncpois()`, and `modtruncpois()` give the mean, variance, median, and mode. `mletruncpois()` estimates the rate parameter (with standard error) from a sample of observed counts via maximum likelihood, and `plottruncpois()` visualizes the PMF, CDF, or quantile function of a truncated Poisson distribution against its non-truncated counterpart.

## Installation

You can install the development version of `truncpois` from GitHub:

```r
# install.packages("pak")
pak::pak("arunsundar022/truncpois")
```

You can then explore the package with:

```r
library(truncpois)
help(dtruncpois)  # Help on the core density function
```

For a more thorough introduction, a vignette document is available as follows:

```r
vignette("truncpois", package = "truncpois")
```

If the vignette is not automatically built when installing from GitHub, include it explicitly:

```r
devtools::install_github("arunsundar022/truncpois", build_vignettes = TRUE)
```

## Quick Example

```{r setup}
set.seed(123)
```

Compare the standard Poisson and right-truncated Poisson PMFs:

```{r pmf-compare}
x <- 0:12
p_pois <- dpois(x, lambda = 5)
p_trunc <- dtruncpois(x, lambda = 5, a = 0, b = 8)

round(data.frame(x = x, p_pois = p_pois, p_trunc = p_trunc), 4)
```

```{r pmf-plot, echo=TRUE}
barplot(rbind(Poisson = p_pois, Truncated = p_trunc),
  beside = TRUE, names.arg = x,
  col = c("grey70", "steelblue"),
  ylab = "Probability", xlab = "x",
  main = "PMF: Poisson(5) vs TruncPois(5, b=8)",
  legend.text = c("Poisson(5)", "TruncPois(5, b=8)")
)
```

Generate random draws from a doubly truncated Poisson distribution:

```{r rtruncpois}
samples <- rtruncpois(1000, lambda = 4, a = 2, b = 9, method = "bounded")
table(samples)
```

Compare empirical and theoretical moments:

```{r moments}
c(
  empirical_mean = mean(samples),
  theoretical_mean = extruncpois(4, a = 2, b = 9)
)

c(
  empirical_var = var(samples),
  theoretical_var = vartruncpois(4, a = 2, b = 9)
)
```

Compute median and mode of truncated distributions:

```{r median-mode}
medtruncpois(lambda = 4, a = 2, b = 9)

# Mode with tied values (lambda = 4 is an integer)
modtruncpois(lambda = 4, a = 2, b = 9)

# Unique mode (lambda = 2.5 is non-integer)
modtruncpois(lambda = 2.5, a = 0, b = 10)
```

## References

Nadarajah, S. and Kotz, S. (2006). R programs for truncated distributions.
_Journal of Statistical Software, Code Snippets_, 16(2), 1--8.
<[doi:10.18637/jss.v016.c02](https://doi.org/10.18637/jss.v016.c02)>
