---
title: "truncpois: Truncated Poisson Distribution"
author: "Arun Sundar Paulraj and Keefe Murphy"
date: "`r Sys.Date()`"
urlcolor: blue
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the truncpois Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{LaplacesDemon}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  fig.show   = "hold",
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE
)
```

## Introduction

The **truncated Poisson distribution** arises when count observations are restricted to a subset of the non-negative integers. Two common cases are the _zero-truncated_ Poisson (counts of at least 1, as in species abundance studies or insurance claims), and the _doubly-truncated_ Poisson (counts confined to an interval $[a, b]$, as in bounded measurement processes).

The __truncpois__ package provides a complete suite of functions for the truncated Poisson distribution, following the naming convention of base R's distribution functions:

| Function | Description |
|---|---|
| `dtruncpois()` | Probability mass function (PMF) |
| `ptruncpois()` | Cumulative distribution function (CDF) |
| `qtruncpois()` | Quantile function (inverse CDF) |
| `rtruncpois()` | Random sample generation |
| `extruncpois()` | Expected value (mean) |
| `vartruncpois()` | Variance |
| `medtruncpois()` | Median |
| `modtruncpois()` | Mode |

All functions accept `a` (lower truncation bound, inclusive, default `0`) and `b` (upper truncation bound, inclusive, default `Inf`). All internal computations are performed on the log-scale for numerical stability, following the framework of Nadarajah and Kotz (2006).

### Installation

```{r install, eval=FALSE}
# Install the development version from GitHub
# install.packages("pak")
pak::pak("arunsundar022/truncpois")
```

```{r load}
library(truncpois)
```

---

## Distribution Functions

### Probability Mass Function

`dtruncpois(x, lambda, a, b)` evaluates the PMF at integer values `x`. With default bounds (`a = 0`, `b = Inf`) it reduces to the standard `dpois()`.

```{r pmf-table}
x          <- 0:12
p_pois     <- stats::dpois(x, lambda = 5)
p_trunc    <- dtruncpois(x, lambda = 5, a = 2, b = 10)

round(rbind(x = x, Poisson = p_pois, Truncated = p_trunc), 4)
```

Notice that `dtruncpois` returns 0 for $x < a = 2$ and $x > b = 10$, and the probabilities over the support $[2, 10]$ are rescaled so that they sum to 1 over the entire support.

```{r pmf-plot, fig.height=4.5}
barplot(
  rbind(Poisson = p_pois, Truncated = p_trunc),
  beside    = TRUE,
  names.arg = x,
  col       = c("grey70", "steelblue"),
  ylab      = "Probability",
  xlab      = "x",
  main      = expression("PMF: Poisson(5) vs TruncPois(5, a=2, b=10)"),
  legend.text = c("Poisson(5)", "TruncPois(5, a=2, b=10)")
)
```

The `log = TRUE` argument returns log-probabilities for numerical work involving very large $\lambda$ or wide truncation windows:

```{r pmf-log}
dtruncpois(100, lambda = 100, a = 50, b = 150, log = TRUE)
```

### Cumulative Distribution Function

`ptruncpois(q, lambda, a, b)` computes $P(X \leq q \mid a \leq X \leq b)$. The CDF reaches exactly 1 at the upper bound $b$ and is 0 for $q < a$.

```{r cdf-table}
q          <- 0:12
cdf_pois   <- stats::ppois(q, lambda = 5)
cdf_trunc  <- ptruncpois(q, lambda = 5, a = 2, b = 10)

round(rbind(q = q, Poisson = cdf_pois, Truncated = cdf_trunc), 4)
```

```{r cdf-plot, fig.height=4.5}
plot(q, cdf_trunc, type = "s", col = "steelblue", lwd = 2,
     ylim = c(0, 1), ylab = "Cumulative Probability", xlab = "q",
     main = "CDF: Poisson(5) vs TruncPois(5, a=2, b=10)")
lines(q, cdf_pois, type = "s", col = "grey50", lwd = 2, lty = 2)
legend("bottomright",
       legend = c("TruncPois(5, a=2, b=10)", "Poisson(5)"),
       col = c("steelblue", "grey50"), lwd = 2, lty = c(1, 2), bty = "n")
```

The `lower.tail` and `log.p` arguments mirror the conventions of `ppois()`:

```{r cdf-uppertail}
# Upper-tail probability P(X > 7 | 2 <= X <= 10)
ptruncpois(7, lambda = 5, a = 2, b = 10, lower.tail = FALSE)
```

### Quantile Function

`qtruncpois(p, lambda, a, b)` is the right-continuous inverse of the CDF: the smallest integer $x$ in $[a, b]$ with $P(X \leq x) \geq p$.

```{r quantile}
p <- seq(0.1, 0.9, by = 0.2)
q <- qtruncpois(p, lambda = 5, a = 2, b = 10)
rbind(p = p, q = q)
```

We can verify the round-trip property — the CDF evaluated at the returned quantiles should be $\geq$ the target probabilities:

```{r quantile-roundtrip}
cdf_at_q <- ptruncpois(q, lambda = 5, a = 2, b = 10)
rbind(target_p = p, quantile = q, cdf_at_q = round(cdf_at_q, 4))
```

---

## Random Generation

`rtruncpois(n, lambda, a, b, method)` draws $n$ independent samples from the truncated Poisson. Three sampling algorithms are available:

- **`"direct"`** (default): Enumerates the support and samples via the Gumbel-max trick. Fast when the support is small or moderate.
- **`"inversion"`**: Inversion sampling via `qtruncpois()`. Works for any parameter configuration.
- **`"bounded"`**: CDF-inversion through the untruncated Poisson. Efficient when truncation removes little probability mass.

```{r random-gen}
set.seed(2026)
x_direct    <- rtruncpois(1000, lambda = 3, a = 1, b = Inf, method = "direct")
x_inversion <- rtruncpois(1000, lambda = 3, a = 1, b = Inf, method = "inversion")
x_bounded   <- rtruncpois(1000, lambda = 3, a = 1, b = Inf, method = "bounded")
```

All three methods produce samples entirely within $[a, b]$ and converge to the same distribution:

```{r random-compare}
rbind(
  direct     = c(min = min(x_direct),    max = max(x_direct),    mean = mean(x_direct)),
  inversion  = c(min = min(x_inversion), max = max(x_inversion), mean = mean(x_inversion)),
  bounded    = c(min = min(x_bounded),   max = max(x_bounded),   mean = mean(x_bounded)),
  theoretical = c(min = NA, max = NA,    mean = extruncpois(3, a = 1))
)
```

For a heavily truncated distribution where the support is wide, use `method = "inversion"` to avoid the finite-support requirement of `"direct"`:

```{r random-large-lambda}
set.seed(42)
samples <- rtruncpois(10, lambda = 100, a = 90, b = 110, method = "inversion")
samples
```

---

## Moments

### Mean

`extruncpois(lambda, a, b)` computes $E[X \mid a \leq X \leq b]$ entirely on the log-scale. For the zero-truncated Poisson ($a = 1$), the mean is always strictly greater than $\lambda$:

```{r mean-ztp}
lambda_val <- 2
c(lambda = lambda_val, "E[X | X >= 1]" = extruncpois(lambda_val, a = 1))
```

As the upper bound $b$ increases toward infinity, the truncated mean converges to $\lambda$:

```{r mean-convergence, fig.height=4.5}
bounds <- 1:15
means  <- sapply(bounds, function(b) extruncpois(lambda = 5, a = 0, b = b))

plot(bounds, means, type = "b", pch = 19, col = "steelblue",
     ylab = "E[X | 0 <= X <= b]", xlab = "Upper Bound b",
     main = "Mean of Right-Truncated Poisson(λ=5) vs Upper Bound")
abline(h = 5, lty = 2, col = "firebrick", lwd = 1.5)
legend("bottomright", legend = c("Truncated mean", "λ = 5"),
       col = c("steelblue", "firebrick"), pch = c(19, NA),
       lty = c(1, 2), bty = "n")
```

### Variance

`vartruncpois(lambda, a, b)` computes the variance via $E[X^2] - (E[X])^2$, keeping all CDF evaluations on the log-scale. Truncation induces _underdispersion_ relative to the standard Poisson (for which $\text{Var}(X) = \lambda$):

```{r variance}
lambda_val <- 5
c("Untruncated variance (= lambda)" = lambda_val,
  "Truncated variance [2, 8]"       = vartruncpois(lambda_val, a = 2, b = 8))
```

We can verify this computationally via a large simulation:

```{r variance-sim}
set.seed(99)
samples <- rtruncpois(20000, lambda = 5, a = 2, b = 8)
c("Simulated variance"   = round(var(samples), 4),
  "Theoretical variance" = round(vartruncpois(5, a = 2, b = 8), 4))
```

### Median

`medtruncpois(lambda, a, b)` returns the median using log-scale inversion of the midpoint CDF value. The median is guaranteed to lie within $[a, b]$:

```{r median}
c("Median (default bounds)" = medtruncpois(lambda = 5),
  "Median [1, Inf]"         = medtruncpois(lambda = 3, a = 1),
  "Median [2, 10]"          = medtruncpois(lambda = 5, a = 2, b = 10),
  "Median (low lambda=0.5)" = medtruncpois(lambda = 0.5, a = 0, b = 5))
```

### Mode

`modtruncpois(lambda, a, b)` returns the most probable value(s) within $[a, b]$. When $\lambda$ is **non-integer**, the unique mode is $\lfloor \lambda \rfloor$. When $\lambda$ is a **positive integer**, both $\lambda - 1$ and $\lambda$ are equally probable, giving tied modes and a warning:

```{r mode-unique}
modtruncpois(lambda = 2.5, a = 0, b = 10) # unique mode: floor(2.5) = 2
```

```{r mode-tie}
modtruncpois(lambda = 5, a = 0, b = 10)   # tied modes at 4 and 5, warning issued
```

When the untruncated mode falls outside $[a, b]$, the mode is clamped to the nearest bound:

```{r mode-clamped}
modtruncpois(lambda = 1, a = 3, b = 10)   # clamped to a = 3
```

---

## Comparison with LaplacesDemon

The `LaplacesDemon` package also provides a truncated Poisson distribution via
`dtrunc()`, `ptrunc()`, `qtrunc()`, and `rtrunc()` with `spec = "pois"`.
This section demonstrates two important limitations of that implementation that
**truncpois** addresses, making it a meaningful advance over existing open-source
software for the truncated Poisson distribution.

```{r laplacesdemon-setup}
ld_available <- requireNamespace("LaplacesDemon", quietly = TRUE)
```

### 1. Incomplete handling of `lower.tail` and `log.p`

`ptruncpois()` and `qtruncpois()` correctly handle all four combinations of
`lower.tail` and `log.p`, computing everything on the log-scale to avoid
numerical cancellation. `LaplacesDemon::ptrunc()` does not support these
arguments — it always returns the lower-tail probability $P(X \leq q)$ on the
linear scale only.

```{r laplacesdemon-lowtail, eval=ld_available}
# Lower-tail probability P(X <= 7 | 2 <= X <= 10, lambda = 5)
ptruncpois(7, lambda = 5, a = 2, b = 10)
LaplacesDemon::ptrunc(7, spec = "pois", a = 2, b = 10, lambda = 5)
```

To obtain an upper-tail probability or a log-probability from `LaplacesDemon`,
users must transform the result manually:

```{r laplacesdemon-uppertail, eval=ld_available}
# Upper-tail probability P(X > 7 | 2 <= X <= 10, lambda = 5)

# truncpois: native lower.tail and log.p support, computed on log-scale
ptruncpois(7, lambda = 5, a = 2, b = 10, lower.tail = FALSE)
ptruncpois(7, lambda = 5, a = 2, b = 10, lower.tail = FALSE, log.p = TRUE)

# LaplacesDemon: manual subtraction and log() required
1 - LaplacesDemon::ptrunc(7, spec = "pois", a = 2, b = 10, lambda = 5)
log(1 - LaplacesDemon::ptrunc(7, spec = "pois", a = 2, b = 10, lambda = 5))
```

This manual approach is error-prone and, for extreme probabilities where
$P(X \leq q)$ is very close to 1, suffers from catastrophic cancellation:
the subtraction $1 - p$ loses all significant digits, producing an
inaccurate log-probability. `truncpois` avoids this entirely by computing
log-probabilities directly using numerically stable `log1p` and `expm1`
arithmetic throughout.

Similarly, `qtruncpois()` accepts `log.p = TRUE` and `lower.tail = FALSE`,
while `LaplacesDemon::qtrunc()` does not:

```{r laplacesdemon-qtrunc, eval=ld_available}
# Upper-tail quantile: smallest x such that P(X > x) <= 0.1
qtruncpois(0.1, lambda = 5, a = 2, b = 10, lower.tail = FALSE)

# Same result using log-scale input
qtruncpois(log(0.1), lambda = 5, a = 2, b = 10, lower.tail = FALSE, log.p = TRUE)

# LaplacesDemon: no lower.tail support -- returns the lower-tail 10th percentile
# instead of the upper-tail 10th percentile
LaplacesDemon::qtrunc(0.1, spec = "pois", a = 2, b = 10, lambda = 5)
```

### 2. No closed-form moments

`LaplacesDemon` provides no functions for the mean, variance, median, or mode of
the truncated Poisson distribution. Users must resort to Monte Carlo simulation,
which is both slow and introduces sampling variability. **truncpois** provides
exact closed-form results for all four moments:

```{r laplacesdemon-moments, eval=ld_available}
# Exact closed-form moments from truncpois (deterministic, no approximation error)
extruncpois(lambda = 5, a = 2, b = 10)
vartruncpois(lambda = 5, a = 2, b = 10)
medtruncpois(lambda = 5, a = 2, b = 10)
suppressWarnings(modtruncpois(lambda = 5, a = 2, b = 10))

# LaplacesDemon: no equivalent -- simulation is the only option
set.seed(42)
samp <- LaplacesDemon::rtrunc(100000, spec = "pois", a = 2, b = 10, lambda = 5)
mean(samp)   # approximation -- varies with sample size and seed
var(samp)    # approximation -- varies with sample size and seed
```

The simulation approach has non-zero sampling error, scales poorly for extreme
truncation parameters, and cannot be used where reproducibility or analytical
precision is required. `truncpois` returns the exact value in closed form for
any $(\lambda, a, b)$.

---

## References

Nadarajah, S. and Kotz, S. (2006). R Programs for Computing Truncated Distributions. _Journal of Statistical Software_, **16**(2). [doi:10.18637/jss.v016.c02](https://doi.org/10.18637/jss.v016.c02)
